---
title: "Modelling Risk Factors and Time Lags in Disease Incidence Prediction"
author: "Joshua Edefo"
date: "2025-02-17"
email: "edefojoshua2000@yahoo.com"
output: github_document
---
The code models the risk of developing coronary heart disease (CHD) by generating a synthetic dataset of 1,000 individuals with varying risk factors such as age, sex, deprivation index, alcohol consumption, smoking, and obesity. It estimates time lags between exposure to risk factors and disease onset using a shifted binomial distribution. The model calculates combined relative risks (RR) for individuals, considering factors like obesity and smoking, and adjusts the CHD incidence based on these RRs. The final dataset provides personalized risk predictions, accounting for time lags and risk factors. This framework aids in predictive health modelling and can inform targeted interventions for disease prevention.

Libraries
```{r a, message=FALSE}
library(dplyr)
library(usethis)
```

Risk modelling
```{r b}
## Step 1: Create a synthetic dataset with age, sex, DIMD, and risk factors
set.seed(123)  # For reproducibility

n <- 1000  ## Number of individuals

## Generate synthetic data
data <- data.frame(
  ID = 1:n,
  Age = sample(30:80, n, replace = TRUE),  ## Random ages between 30-80
  Sex = sample(c("Male", "Female"), n, replace = TRUE),
  DIMD = sample(1:5, n, replace = TRUE),  ## Deprivation index (1 = least deprived, 5 = most deprived)
  Alcohol_Intake = rbinom(n, 1, 0.3),  ## 30% of individuals consume alcohol
  Smoking = rbinom(n, 1, 0.25),  ## 25% are smokers
  Obesity = rbinom(n, 1, 0.35)  ## 35% are obese
)

## Step 2: Define a function to estimate time lags based on a shifted binomial distribution
estimate_time_lag <- function(mean_lag, min_lag = 1, max_lag = 10) {
  time_lag <- rbinom(1, max_lag - min_lag, mean_lag / max_lag) + min_lag
  return(time_lag)
}

## Step 3: Assign time lags for each individual based on risk-disease pairs
data <- data %>%
  mutate(
    Alcohol_BC_Lag = ifelse(Alcohol_Intake == 1, estimate_time_lag(9), NA),  ## 9-year mean lag for alcohol and breast cancer
    Smoking_CHD_Lag = ifelse(Smoking == 1, estimate_time_lag(5), NA),  ## 5-year mean lag for smoking and CHD
    Obesity_CHD_Lag = ifelse(Obesity == 1, estimate_time_lag(7), NA)  ## 7-year mean lag for obesity and CHD
  )

## Step 4: Compute incidence probabilities based on the multiplicative risk model

## Relative risks for CHD:
RR_Obesity_CHD <- 1.5   # Relative risk for obesity and coronary heart disease
RR_Smoking_CHD <- 1.3   # Relative risk for smoking and coronary heart disease

## Compute combined relative risk for each individual
data <- data %>%
  mutate(
    Combined_RR_CHD = case_when(
      Obesity == 1 & Smoking == 1 ~ RR_Obesity_CHD * RR_Smoking_CHD,  # Obese smoker
      Obesity == 1 & Smoking == 0 ~ RR_Obesity_CHD,  # Obese non-smoker
      Smoking == 1 & Obesity == 0 ~ RR_Smoking_CHD,  # Smoker non-obese
      TRUE ~ 1  # Non-smokers and non-obese individuals
    )
  )

## Step 5: Calculate incidence based on the combined relative risk (adjusted for 2013 baseline incidence)
## Example total incidence of CHD for 2013 (let's assume 0.05 or 5%)
total_incidence_2013 <- 0.05

## Adjusted incidence based on risk factors
data <- data %>%
  mutate(
    Adjusted_Incidence_CHD = total_incidence_2013 * Combined_RR_CHD
  )

## View the first few rows of the modified data
head(data)


```

session information

```{r c, echo=FALSE}
sessionInfo()

```
